Data Processing

### 2025-04-17 code/processing-code/01_data_cleaning.qmd
Few minor changes

---
title: "Final project"
author: "Murtaza Yaqubi"
format: html
editor: visual
---

# Load necessary libraries

```{r}
library(knitr)
library(readr)
library(dplyr)
library(ggplot2)
library(ggcorrplot)
library(vcd)
library(caret)        # For data partitioning and model training
library(randomForest) # For Random Forest modeling
library(tidyverse)
library(tidymodels)
library(broom)
library(here)           # For file paths
library(pROC)
library(naniar)         # For gg_miss_var()
```

# PART 1: DATA IMPORT AND CLEANING

```{r}
# Import the dataset
cardio_df <- read_csv(here("data", "raw-data", "Cardiovascular_Disease_Dataset.csv"))

@@

# PART 2: EXPLORATORY DATA ANALYSIS (EDA)

```{r}
# Calculate summary statistics for selected variables
summary_stats <- cardio_df %>% 
  summarise(
    Age_min = min(age, na.rm = TRUE),
    Age_max = max(age, na.rm = TRUE),
    Age_mean = mean(age, na.rm = TRUE),
    Age_sd = sd(age, na.rm = TRUE),

    RestingBP_min = min(restingBP, na.rm = TRUE),
    RestingBP_max = max(restingBP, na.rm = TRUE),
    RestingBP_mean = mean(restingBP, na.rm = TRUE),
    RestingBP_sd = sd(restingBP, na.rm = TRUE),

    SerumCholestrol_min = min(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
    SerumCholestrol_max = max(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
    SerumCholestrol_mean = mean(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
    SerumCholestrol_sd = sd(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),

    MaxHeartRate_min = min(maxheartrate, na.rm = TRUE),
    MaxHeartRate_max = max(maxheartrate, na.rm = TRUE),
    MaxHeartRate_mean = mean(maxheartrate, na.rm = TRUE),
    MaxHeartRate_sd = sd(maxheartrate, na.rm = TRUE),

    Oldpeak_min = min(oldpeak[oldpeak > 0], na.rm = TRUE),
    Oldpeak_max = max(oldpeak, na.rm = TRUE),
    Oldpeak_mean = mean(oldpeak, na.rm = TRUE),
    Oldpeak_sd = sd(oldpeak, na.rm = TRUE)
  )

# Reshape into long format, then wide table
summary_table_formatted <- summary_stats %>%
  pivot_longer(cols = everything(), names_to = "stat", values_to = "value") %>%
  separate(stat, into = c("Variable", "Statistic"), sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = Variable, values_from = value)

# View formatted table
summary_table_formatted

# save to file
summarytable_file = here("results", "tables", "table1_summary_statistics.rds")
saveRDS(summary_table_formatted, file = summarytable_file)

# Age distribution
p1 <- ggplot(cardio_df, aes(x = age)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "steelblue", alpha = 0.5) +
  geom_density(color = "darkblue", linewidth = 1) +
  labs(title = "Age Distribution", x = "Age", y = "Density") +
  scale_x_continuous(breaks = seq(0, max(cardio_df$age, na.rm = TRUE) + 10, by = 10)) +
  theme_bw()

plot(p1)
figure_file = here("results", "figures","Age distribution.png")
ggsave(filename = figure_file, plot=p1)

# Serum Cholesterol distribution (excluding zeros)
p2 <- ggplot(subset(cardio_df, serumcholestrol > 0), aes(x = serumcholestrol)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "forestgreen", alpha = 0.5) +
  geom_density(color = "darkgreen", linewidth = 1) +
  labs(title = "Serum Cholesterol Distribution", 
       x = "Serum Cholesterol (mg/dL)", y = "Density") +
  theme_minimal()

plot(p2)
figure_file = here("results", "figures","Serum Cholestrol.png")
ggsave(filename = figure_file, plot=p2)

# Resting Blood Pressure distribution (excluding zeros)
p3 <- ggplot(subset(cardio_df, restingBP > 0), aes(x = restingBP)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "maroon", alpha = 0.5) +
  geom_density(color = "darkred", linewidth = 1) +
  labs(title = "Resting Blood Pressure Distribution", 
       x = "Resting Blood Pressure (mmHG)", y = "Density") +
  theme_bw()

plot(p3)
figure_file = here("results", "figures","Resting Blood Pressure Distribution.png")
ggsave(filename = figure_file, plot=p3)

# Boxplot for resting blood pressure by heart disease status
p4 <- ggplot(cardio_df, aes(x = target, y = restingBP, fill = target)) +
  geom_boxplot() +
  labs(title = "Resting Blood Pressure by Heart Disease Status",
       x = "Heart Disease Status",
       y = "Resting Blood Pressure (mmHg)") +
  theme_minimal()

plot(p4)
figure_file = here("results", "figures","restingBP by heart disease status.png")
ggsave(filename = figure_file, plot=p4)

# Violin plot for oldpeak by heart disease status
p5 <- ggplot(cardio_df, aes(x = target, y = oldpeak, fill = target)) +
  geom_violin(trim = FALSE) +
  labs(title = "Oldpeak Distribution by Heart Disease Status",
       x = "Heart Disease Status",
       y = "Oldpeak") +
  theme_minimal()

plot(p5)
figure_file = here("results", "figures","oldpeak by heart disease status.png")
ggsave(filename = figure_file, plot=p5)

# Create the correlation matrix for selected numeric variables
numeric_vars <- cardio_df %>% 
  select(age, restingBP, serumcholestrol, maxheartrate, oldpeak)
cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")

# Plot the correlation matrix with ggcorrplot
p6 <- ggcorrplot(cor_matrix, 
           lab = TRUE,                         
           lab_size = 3,                       
           colors = c("blue", "white", "red"),  # Color gradient from blue (negative) to red (positive)
           title = "Correlation Matrix", 
           ggtheme = theme_minimal())

plot(p6)
figure_file = here("results", "figures","correlation matrix.png")
ggsave(filename = figure_file, plot=p6)     
```

# PART 3: STATISTICAL ANALYSIS: DATA CLEANING, SUMMARIZATION, AND VISUALIZATION

```{r}
# Filter the data for individuals under 50 with normal cholesterol (<200 mg/dL) and create risk indicator variables for elevated blood pressure and ST depression.
cleaned_df <- cardio_df %>%
  filter(age < 50, serumcholestrol > 0, serumcholestrol < 200) %>%  
  mutate(
    # Create a binary variable for blood pressure:
    # "Elevated" if restingBP is between 120 and 129; otherwise, "Otherwise"
    elevated_BP = if_else(between(restingBP, 120, 129), "Elevated", "Otherwise"),

    # Create a binary indicator for ST depression:
    # "Elevated" if oldpeak is greater than 1; otherwise, "Normal"
    elevated_oldpeak = if_else(oldpeak > 1, "Elevated", "Normal")
  ) %>% 
  # Select the variables we want to keep
  select(age, gender, chestpain, restingBP, serumcholestrol, fastingbloodsugar, 
         maxheartrate, oldpeak, target, elevated_BP, elevated_oldpeak)

# Print a sample of the cleaned dataset
head(cleaned_df)

# Save the processed data
saveRDS(cleaned_df, file = "../../data/processed-data/cleaned_df.rds")

# Create a summary table by grouping the cleaned data by blood pressure status, ST depression status, and heart disease (target)
summary_table <- cleaned_df %>%
  group_by(elevated_BP, elevated_oldpeak, target) %>%
  summarise(count = n(), .groups = "drop")
print(summary_table)

summary_df = skimr::skim(summary_table)
print(summary_table)
# save to file
summarytable_file = here("results", "tables", "summarytable2.rds")
saveRDS(summary_df, file = summarytable_file)

# Interaction Visualization: Heatmap
interaction_counts <- cleaned_df %>%
  group_by(elevated_BP, elevated_oldpeak, target) %>%
  summarise(count = n(), .groups = "drop")

p7 <- heatmap_plot <- ggplot(interaction_counts, aes(x = elevated_BP, y = elevated_oldpeak, fill = count)) +
  geom_tile(color = "white", size = 0.5) +
  geom_text(aes(label = count), color = "black", size = 5) +
  facet_wrap(~ target, ncol = 2) +
  scale_fill_gradient(low = "lightgreen", high = "forestgreen") +
  labs(title = "Interaction between Blood Pressure and ST Depression",
       subtitle = "By Heart Disease Status (Individuals <50 with Normal Cholesterol)",
       x = "Blood Pressure Status",
       y = "ST Depression Status",
       fill = "Count") +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

plot(p7)
figure_file = here("results", "figures","interaction heatmap.png")
ggsave(filename = figure_file, plot=p7)

# Visualize CVD risk by blood pressure and ST depression status using faceted grouped Bar Chart
p8 <- ggplot(interaction_counts, aes(x = elevated_BP, y = count, fill = target)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
  facet_wrap(~ elevated_oldpeak, ncol = 2, scales = "free_y") +
  labs(title = "CVD Risk by Blood Pressure and ST Depression Status",
       subtitle = "Individuals Under 50 with Normal Cholesterol (<200 mg/dL)",
       x = "Blood Pressure Status",
       y = "Count",
       fill = "Heart Disease") +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  )

plot(p8)
figure_file = here("results", "figures","CVD risk by BP and ST depression.png")
ggsave(filename = figure_file, plot=p8)
```

# PART 4a: INFERENTIAL STATISTICAL ANALYSIS

# Chi-Square Tests

```{r}
# Create a contingency table for blood pressure status vs. heart disease
table_bp_target <- table(cleaned_df$elevated_BP, cleaned_df$target)
print(table_bp_target)
chisq_test_bp <- chisq.test(table_bp_target)
print(chisq_test_bp)

# Create a contingency table for ST depression vs. heart disease
table_oldpeak_target <- table(cleaned_df$elevated_oldpeak, cleaned_df$target)
print(table_oldpeak_target)
chisq_test_oldpeak <- chisq.test(table_oldpeak_target)
print(chisq_test_oldpeak)
```

# Logistic Regression Analysis

```{r}
# Convert target to a binary outcome for regression (1 = Heart Disease, 0 = No Heart Disease)
cleaned_df <- cleaned_df %>%
  mutate(target_binary = if_else(target == "Heart Disease", 1, 0))

# Fit a logistic regression model with elevated_BP and elevated_oldpeak as predictors
model <- glm(target_binary ~ elevated_BP + elevated_oldpeak, 
             data = cleaned_df, family = binomial)
summary(model)

# save fit results table  
table_file1 = here("results", "tables", "resulttable1.rds")
saveRDS(model, file = table_file1)

# Fit a logistic regression model including the interaction term
model_interaction <- glm(target_binary ~ elevated_BP * elevated_oldpeak, 
                         data = cleaned_df, family = binomial)
summary(model_interaction)

# save fit results table  
table_file1 = here("results", "tables", "resulttable2.rds")
saveRDS(model_interaction, file = table_file1)
```

# Extract Odds Ratios (OR) from the logistic regression model with interaction

```{r}
or_values <- exp(coef(model_interaction))
or_ci <- exp(confint(model_interaction))
or_results <- data.frame(
  Term = names(or_values),
  OR = or_values,
  CI_lower = or_ci[, 1],
  CI_upper = or_ci[, 2]
)
print("Odds Ratios (with 95% CI):")
print(or_results)

# save fit results table  
table_file2 = here("results", "tables", "table2_odds_ratios.rds")
saveRDS(or_results, file = table_file2)

# Tidy the model output and compute odds ratios and 95% CI
tidy_model <- tidy(model_interaction, exponentiate = TRUE, conf.int = TRUE)
print(tidy_model)
```

# Fit a log-binomial model to estimate Relative Risks (RR)

```{r}
rr_model <- glm(target_binary ~ elevated_BP + elevated_oldpeak, 
                data = cleaned_df, family = binomial(link = "log"))
rr_values <- exp(coef(rr_model))
rr_ci <- exp(confint(rr_model))
rr_results <- data.frame(
  Term = names(rr_values),
  RR = rr_values,
  CI_lower = rr_ci[, 1],
  CI_upper = rr_ci[, 2]
)
print("Relative Risks (with 95% CI):")
print(rr_results)

# save fit results table  
table_file3 = here("results", "tables", "table3_relative_risks.rds")
saveRDS(rr_results, file = table_file3)
```

# ROC Curve and AUC for the logistic regression model with interaction

```{r}
roc_curve <- roc(cleaned_df$target_binary, predict(model_interaction, type = "response"))
plot(roc_curve, main = "ROC Curve for Logistic Regression Model", col = "blue", lwd = 2)
auc_value <- auc(roc_curve)
print(paste("AUC:", auc_value))

# Save ROC plot
roc_fig <- recordPlot()
saveRDS(roc_fig, file = here("results", "figures", "figure1_roc_logistic_model.rds"))

```

# Adjusted Logistic Regression Analysis including confounders (gender and fasting blood sugar)

```{r}
adjusted_model <- glm(target_binary ~ elevated_BP * elevated_oldpeak + gender + fastingbloodsugar, 
                      data = cleaned_df, family = binomial)
summary(adjusted_model)

# Extract Odds Ratios (OR) for the adjusted model
adj_or_values <- exp(coef(adjusted_model))
adj_or_ci <- exp(confint(adjusted_model))
adj_or_results <- data.frame(
  Term = names(adj_or_values),
  OR = adj_or_values,
  CI_lower = adj_or_ci[, 1],
  CI_upper = adj_or_ci[, 2]
)
print("Adjusted Odds Ratios (with 95% CI):")
print(adj_or_results)

# save fit results table  
table_file4 = here("results", "tables", "table4_adjusted_odds_ratios.rds")
saveRDS(adj_or_results, file = table_file4)

# Tidy the adjusted model output and compute odds ratios and 95% CI
tidy_adjusted_model <- tidy(adjusted_model, exponentiate = TRUE, conf.int = TRUE)
print(tidy_adjusted_model)
```

# PART 4b: MODEL BUILDING

```{r}
# Set seed for reproducibility
rdmseed <- 123

# Reset seed before modeling
set.seed(rdmseed)

cleaned_df <- cleaned_df %>%
  mutate(target_binary = if_else(target == "Heart Disease", 1, 0))
```

# linear model

```{r}
# Fit a standard linear regression model using predictors
lm_model <- linear_reg() %>%
  fit(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df)

# Get predictions from linear model and calculate RMSE
cleaned_df$pred_lm <- predict(lm_model, new_data = cleaned_df)$.pred
rmse(cleaned_df, truth = target_binary, estimate = pred_lm)

# # Plot observed vs. predicted for Linear Model
ggplot(cleaned_df, aes(x = target_binary, y = pred_lm)) +
  geom_point(color = "darkblue", alpha = 0.7, size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))

```

# LASSO

```{r}
set.seed(rdmseed)

# Prepare recipe for LASSO with normalization
rec <- recipe(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df) %>%
  step_dummy(elevated_BP, elevated_oldpeak) %>%   # Convert categorical predictors to dummies
  step_normalize(all_numeric(), -all_outcomes())

# Define LASSO model with a fixed penalty
lasso_spec <- linear_reg(penalty = 0.1, mixture = 1) %>%
  set_engine("glmnet")

# Create a workflow and fit the LASSO model
wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(lasso_spec)

lasso_fit <- fit(wf, data = cleaned_df)
cleaned_df$pred_lasso <- predict(lasso_fit, new_data = cleaned_df)$.pred
rmse(cleaned_df, truth = target_binary, estimate = pred_lasso)

# Plot observed vs. predicted for LASSO Model
ggplot(cleaned_df, aes(x = target_binary, y = pred_lasso)) +
  geom_point(color = "darkviolet", alpha = 0.7, size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))
```

# Random Forest

```{r}
set.seed(rdmseed)

# Create a workflow and fit the Random Forest model
rf_model <- rand_forest() %>%
  set_mode("regression") %>%
  set_engine("ranger", seed = rdmseed)

rf_fit <- rf_model %>%
  fit(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df)

# Get predictions and evaluate model performance
cleaned_df$pred_rf <- predict(rf_fit, cleaned_df)$.pred
rmse(cleaned_df, truth = target_binary, estimate = pred_rf)

# Save RMSEs from all models to a table
rmse_df <- tibble(
  Model = c("Linear", "LASSO", "Random Forest"),
  RMSE = c(
    rmse(cleaned_df, truth = target_binary, estimate = pred_lm)$.estimate,
    rmse(cleaned_df, truth = target_binary, estimate = pred_lasso)$.estimate,
    rmse(cleaned_df, truth = target_binary, estimate = pred_rf)$.estimate
  )
)
saveRDS(rmse_df, file = here("results", "tables", "table5_model_rmse_comparison.rds"))

# Plot observed vs. predicted for RF Model
ggplot(cleaned_df, aes(x = target_binary, y = pred_rf)) +
  geom_point(color = "darkgreen", alpha = 0.7, size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1))

```

# Tuning the LASSO model (without cross-validation)

```{r}
set.seed(rdmseed)

# Create a grid of penalty values for LASSO
lasso_grid <- tibble(penalty = 10^seq(-5, 2, length.out = 50))

# Define LASSO model with tunable penalty
lasso_spec_tune <- linear_reg(penalty = tune()) %>%
  set_engine("glmnet")

# Create workflow and assign to lasso_tuning_workflow (new name)
lasso_tuning_workflow <- workflow() %>%
  add_model(lasso_spec_tune) %>%
  add_recipe(rec)

# Perform grid search without CV (apparent resampling)
lasso_tune <- tune_grid(
  lasso_tuning_workflow,
  resamples = apparent(cleaned_df),
  grid = lasso_grid,
  metrics = metric_set(rmse)
)

# Extract tuning results
lasso_tune_df <- as.data.frame(lasso_tune$.metrics)

# Identify best LASSO penalty
best_lasso <- lasso_tune_df %>% filter(.estimate == min(.estimate))
print("Best LASSO Penalty (No CV):")
print(best_lasso)

# Plot LASSO tuning results
ggplot(lasso_tune_df, aes(x = penalty, y = .estimate)) +
  geom_line(size = 1, color = "steelblue") +
  geom_point(size = 2, color = "firebrick") +
  scale_x_log10() +
  labs(
    x = "Penalty (log scale)",
    y = "RMSE",
    title = "LASSO Tuning (No Cross-validation)",
    subtitle = "Using apparent resampling"
  ) +
  theme_minimal()

# Save LASSO tuning plot
lasso_tune_plot <- last_plot()
ggsave(filename = here("results", "figures", "figure2_lasso_tuning_plot.png"), plot = lasso_tune_plot)
```

# Tuning for our Random Forest model (without cross-validation)

```{r}
# Create tuning grid for mtry and min_n
rf_grid <- grid_regular(
  mtry(range = c(1, 7)),
  min_n(range = c(1, 21)),
  levels = 7
)

# Define random forest model with tunable parameters
rf_spec_tune <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 300
) %>%
  set_mode("regression") %>%
  set_engine("ranger", seed = rdmseed)

# Build workflow
rf_wf_tune <- workflow() %>%
  add_model(rf_spec_tune) %>%
  add_recipe(rec)

# Perform RF tuning without CV
rf_tune <- tune_grid(
  rf_wf_tune,
  resamples = apparent(cleaned_df),
  grid = rf_grid,
  metrics = metric_set(rmse)
)

rf_tune_df <- as.data.frame(rf_tune$.metrics)

# Identify best RF tuning parameters
best_rf <- rf_tune_df %>% filter(.estimate == min(.estimate))
print("Best Random Forest Parameters (No CV):")
print(best_rf)

# Plot RF tuning results
ggplot(rf_tune_df, aes(x = mtry, y = min_n, fill = .estimate)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    title = "Random Forest Tuning (No Cross-validation)",
    x = "mtry",
    y = "min_n",
    fill = "RMSE"
  ) +
  theme_minimal()

# Save Random Forest tuning heatmap
rf_tune_plot <- last_plot()
ggsave(filename = here("results", "figures", "figure3_rf_tuning_heatmap.png"), plot = rf_tune_plot)

```

# Cross-validation for our LASSO model tuning

```{r}
set.seed(rdmseed)

# Create 5-fold CV with 5 repeats
cv_folds <- vfold_cv(cleaned_df, v = 5, repeats = 5)

# Tune LASSO using CV
lasso_tune_cv <- tune_grid(
  lasso_tuning_workflow,
  resamples = cv_folds,
  grid = lasso_grid,
  metrics = metric_set(rmse)
)

autoplot(lasso_tune_cv)
```

# Cross-validation tuning for our random forest model

```{r}
# Tune RF using CV
rf_tune_cv <- tune_grid(
  rf_wf_tune,
  resamples = cv_folds,
  grid = rf_grid,
  metrics = metric_set(rmse)
)
autoplot(rf_tune_cv)

```

>>>>

---
title: "Data Processing"
author: "Murtaza Yaqubi"
format:
  html:
    code-fold: true      # Makes code chunks collapsible
    code-tools: true     # Adds copy/download buttons
    echo: true           # THIS is what ensures code is shown
editor: visual
---

```{r}
# Load necessary libraries
library(knitr)
library(readr)
library(dplyr)
library(ggplot2)
library(ggcorrplot)
library(vcd)
library(caret)        # For data partitioning and model training
library(randomForest) # For Random Forest modeling
library(tidyverse)
library(tidymodels)
library(broom)
library(here)           # For file paths
library(pROC)
library(naniar)         # For gg_miss_var()

# Import the dataset
cardio_df <- read_csv(here("data", "raw-data", "Cardiovascular_Disease_Dataset.csv"))

@@
